### 2021-09-16 README.Rmd
Revisions after decision by Geographical Analysis

The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research. Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception. Relatively few papers share code and data sufficiently, which hinders not only verification but additional experimentation. In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19. Transparency and openness means that independent researchers can, with relatively modest efforts, verify findings and use different approaches as appropriate. Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.

@@

There are various reasons why a project can fail to be reproducible. In some cases, there might be legitimate reasons to withhold the data, perhaps due to confidentiality and privacy reasons [e.g., @Lee2020human]. But in many other cases the data are publicly available, which in fact has commonly been the case with population-level COVID-19 information. Typically the provenance of the data is documented, but in numerous studies the data themselves are not shared [@Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid]. As any researcher can attest, whether a graduate student or a seasoned scientist, collecting, organizing, and preparing data for a project can take a substantial amount of time. Pointing to the sources of data, even when these sources are public, is a small step towards reproducibility-but only a very small one. Faced with the prospect of having to recreate a data set from raw sources is probably sufficient to dissuade all but the most dedicated (or stubborn) researcher from independent verification. This is true even if part of the data are shared [e.g., @Wong2020spreading]. In other cases, data are shared, but the processes followed in the preparation of the data are not fully documented [@Ahmad2020association; @Skorka2020macroecology]. These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [@Herndon2014high] and the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions^[\url{https://doi.org/10.1038/d41586-020-00287-y}]. Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [@Noury2021how; @Pequeno2020air; @Wang2021transmission]. Making code available only "on demand" [e.g., @Brandtner2021creatures] is an unnecessary barrier when most journals offer the facility to share supplemental materials online. Then there are those papers that more closely comply with reproducibility standards, and share well-documented processes and data, as well as the code used in any analyses reported [@Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state; @Sy2021population]. Even in this case, the pressure to publish "new findings" instead of replication studies can act as a deterrent, perhaps particularly for younger researchers^[The present paper was desk rejected by three journals that had previously published research on population density and the spread of COVID-19; in one case, the paper was too opinionated for the journal, in the other two cases, the paper was not a "good fit" despite dealing with a nearly identical issue as papers previously published in said journals. This does not inspire much confidence in the commitment of journals to reproducibility in research.].

@@

In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of an independent researcher with expertise in spatial analysis are asked, and the data are reanalyzed.

@@

SWN examined the association between the basic reproductive number of COVID-19 and population density. The basic reproductive number $R_0$ is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness. In rough terms, $R_0$ measures how many new infections each infections begets. Infectious disease outbreaks generally tend to die out when $R_0<1$, and to grow when $R_0>1$. Reliable calculation of $R_0$ requires a minimum number of cases to be able to assume that there is community transmission of the pathogen. Accordingly, SWN based their analysis only on counties that had at least 25 cases or more at the end of the exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample included `r prettyNum(nrow(county_geo %>% filter(!is.na(R))), big.mark = ",")` counties in the US, including in Alaska, Hawaii, Puerto Rico, and island territories.

```{r R0-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + # Create a ggplot
  # Choropleth map of basic reproductive number
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = R),
          color = NA) +
  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_viridis_c(name = expression(R[0]),
                       direction = -1) +
  # Overlay counties with missing values of the reproductive number in white
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   R == 0), 
          fill = "white",
          color = NA) +
  # Overlay county boundaries
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          fill = NA,
          color = "lightgray") +
  # Add boundary of continental US
  geom_sf(data = us_geo,
          fill = NA) + 
  # Add caption
  labs(caption = "Note: counties in white represent missing values of the basic reproductive number") +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

Table \ref{tab:swn-results} reproduces the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SWN). It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is changed to one percent units, instead of the ten percent units used by SWN). The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fourth row of Table \ref{tab:swn-results}. It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tend to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tend to have higher values of $R_0$. The key of the analysis, after extensive sensitivity analysis, is a robust finding that population density has a positive association with the basic reproductive number. But does it?

@@

```{r tabulate-swn-results, echo=FALSE}
results_swn %>%
  mutate(Variable = c("Intercept", 
                      "Log of population density",
                      "Percent of private transportation", 
                      "Median household income ($10,000)",
                      "Standard deviation (Intercept)",
                      "Within-group standard error")) %>%
  kable("latex",
        digits = 3,
        booktabs = TRUE,
        col.names = c("Variable",
                      "beta",
                      "95% CI",
                      "beta",
                      "95% CI",
                      "beta",
                      "95% CI"),
        caption = "\\label{tab:swn-results}Reproducing SWN: Models 1-3") %>%
  kable_styling(latex_options = c("scale_down")) %>%
  add_header_above(c(" ", "Model 1" = 2, "Model 2" = 2, "Model 3" = 2))
```

```{r random-terms-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + 
  geom_sf(data = state_geo %>%
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = random_intercepts), # Add the simple features object to the plot
          color = "lightgray") +
  # Add boundary of continental US
  geom_sf(data = us_geo,
          fill = NA) + 
  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_viridis(name = "Random Intercepts",
                     option = "turbo",
                     direction = -1) +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r urban-areas-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + # Create a ggplot
  # Choropleth map of basic reproductive number
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = R),
          color = NA) +
  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_viridis_c(name = "R0",
                       direction = -1) +
  # Overlay counties with missing values of the reproductive number in white
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   R == 0), 
          fill = "white",
          color = NA) +
  # Overlay boundaries of urban areas
  geom_sf(data = urban_geo %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   UATYP10 == "U"), # Include only urbanized areas with population > 50,000
          fill = NA,
          color = "red") +
  # Add boundary of continental US
  geom_sf(data = us_geo,
          fill = NA) + 
  # Add caption
  labs(caption = "Note: boundaris of urbanized areas with population > 50,000 are shown in red") +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r tabulate-sample-selection-results, echo=FALSE}
results_selection  %>%
  mutate(Variable = c("Intercept",
                      "Log of population density",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (10,000)",
                      "Intercept",
                      "Log of population density",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (\\$10,000)",
                      "$\\sigma$",
                      "$\\rho$")) %>%
  kable("latex",
        digits = 3,
        booktabs = TRUE,
        escape = FALSE,
        col.names = c("Variable",
                      "$\\beta$", 
                      "95\\% CI",
                      "$\\beta$",
                      "95\\% CI"),
        align = c("lcccc"),
        caption = "\\label{tab:selection-results}Estimation results of sample selection models") %>%
  kable_styling(latex_options = c("scale_down")) %>% 
  pack_rows(group_label = "Sample Selection Model", 1, 6) %>%
  pack_rows(group_label = "Outcome Model", 7, 12) %>%
  add_header_above(c(" ", "Selection Model 1" = 2, "Selection Model 2" = 2))
```

@@

```{r tabulate-sample-selection3-results, echo=FALSE}
selection3_me  %>%
  mutate(Variable = c("Intercept",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (10,000)",
                      "Intercept",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (\\$10,000)",
                      "Spatial filter",
                      "$\\sigma$",
                      "$\\rho$")) %>%
  kable("latex",
        digits = 3,
        booktabs = TRUE,
        escape = FALSE,
        col.names = c("Variable",
                      "$\\beta$", 
                      "95\\% CI"),
        align = c("lcccc"),
        caption = "\\label{tab:selection3-results}Estimation results of sample selection model with spatial filter") %>%
  kable_styling(font_size = 8, latex_options = "hold_position") %>% 
  pack_rows(group_label = "Sample Selection Model", 1, 5) %>%
  pack_rows(group_label = "Outcome Model", 6, 11) %>%
  add_header_above(c(" ", "Selection Model 3" = 2))
```

@@

How relevant are the differences between the various model specifications presented above? Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3. The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1. The conclusion is that at higher densities, $R_0$ is _always_ higher. The two panels on the right, in contrast, shows that Selection Model 2 and Selection Model 3 coincide that $R_0$ tends to increase as density grows. This continues until a density of approximately 2.9 (1,000 people per sq.km). At higher densities than that the relationship between density and $R_0$ begins to weaken, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).

It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p. 201]. There are almost always multiple ways to approach a modelling situation. In the present case, we would argue that spatial sampling is an important aspect of the modeling process, but one that perhaps required different technical skills than those available to SWN. There is nothing wrong with that. What matters is that, by adopting relatively high reproducibility standards, these researchers made a valuable and honest contribution to the collective enterprise of seeking knowledge. Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.

In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [as put by Wong and Li, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading]. At the same time, this also stands in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with it in a longitudinal setting [@Hamidi2020longitudinal. In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al. [-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties. More generally, there is no doubt that in population-level studies density is indicative of proximity, but it also potentially is a proxy for adaptive behavior. And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.

@@

Taking one recent example from the literature [Sy et al., @Sy2021population; SWN], the present paper illustrates the importance of good reproducibility practices. Sharing data and code can catalyze research, by allowing independent verification of findings, as well as additional research. After verifying the results of SWN, experiments with sample selection models and variations in the definition of model inputs, lead to an important reappraisal of the conclusion that high density is associated with greater spread of the virus. Instead, the possibility of a non-monotonical relationship between population density and contagion is raised.

In the spirit of openness, this paper is prepared as an R Markdown document, an a companion data package is provided. The data package contains the relevant documentation of the data, and all data pre-processing is fully documented. Hopefully this, and similar reproducible papers, will continue to encourage others to adopt reproducible standards in their research.

>>>>

The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic in 2019 led to explosive growth in scientific research. Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception. Relatively few papers share code and data sufficiently, which hinders not only verification but additional experimentation. In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19. Transparency and openness means that independent researchers can, with only modest efforts, verify findings and use different approaches as appropriate. Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.

@@

There are various reasons why a project can fail to be reproducible. In some cases, there might be legitimate reasons to withhold the data, perhaps due to confidentiality and privacy reasons [e.g., @Lee2020human]. But in many other cases the data are publicly available, which in fact has commonly been the case with population-level COVID-19 information. Typically the provenance of the data is documented, but in numerous studies the data themselves are not shared [@Amadu2021assessing; @Bhadra2021impact; @Cruz2020exploring; @Feng2020spread; @Fielding2020social; @Hamidi2020longitudinal; @Hamidi2020density; @Inbaraj2021seroprevalence; @Souris2020covid]. As any researcher can attest, collecting, organizing, and preparing data for a project can take a substantial amount of time. Pointing to the sources of data, even when these sources are public, is a small step towards reproducibility-but only a very small one. Faced with the prospect of having to recreate a data set from raw sources is probably sufficient to dissuade all but the most dedicated (or stubborn) researcher from independent verification. This is true even if part of the data are shared [e.g., @Wong2020spreading]. In other cases, data are shared, but the processes followed in the preparation of the data are not fully documented [@Ahmad2020association; @Skorka2020macroecology]. These processes matter, as shown by the errors in the spreadsheets of Reinhart and Rogoff [see @Herndon2014high for the discovery of these errors], as well as by the data of biologist Jonathan Pruitt that led to an "avalanche" of paper retractions [see @Viglione2020avalanche]. Another situation is when papers share well-documented data, but fail to provide the code used in the analysis [@Noury2021how; @Pequeno2020air; @Wang2021transmission]. Making code available only "on demand" [e.g., @Brandtner2021creatures] is an unnecessary barrier when most journals offer the facility to share supplemental materials online. Then there are those papers that more closely comply with reproducibility standards, and share well-documented processes and data, as well as the code used in any analyses reported [@Paez2020spatio; @Feyman2020effectiveness; @Stephens2021impact; @White2020state; @Sy2021population]. Even in this case, the pressure to publish "new findings" instead of replication studies can act as a deterrent, perhaps particularly for younger researchers^[The present paper was desk rejected by three journals that had previously published research on population density and the spread of COVID-19; in one case, the paper was too opinionated for the journal, in the other two cases, the paper was not a "good fit" despite dealing with a nearly identical issue as papers previously published in said journals.].

@@

In the following sections, the analysis of SWN is reproduced, some relevant questions from the perspective of an independent researcher with expertise in spatial analysis are asked, and the data are reanalyzed.

@@

SWN examined the association between the basic reproductive number of COVID-19 and population density. The basic reproductive number $R_0$ is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness. In rough terms, $R_0$ measures how many new infections each infections begets. Infectious disease outbreaks generally tend to die out when $R_0<1$, and to grow when $R_0>1$. Reliable calculation of $R_0$ requires a minimum number of cases to be able to assume that there is community transmission of the pathogen. Accordingly, SWN based their analysis only on counties that had at least 25 cases or more at the end of the exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample included `r prettyNum(nrow(county_geo %>% filter(!is.na(R))), big.mark = ",")` counties in the US, including in Alaska, Hawaii, Puerto Rico, and island territories. SWN used COVID-19 data collected by the New York Times and made available (with versioning) in a GitHub repository^[https://github.com/nytimes/covid-19-data]. For each county, SWN assumed that the exponential growth period began one week prior to the second daily increase in cases, and assumed that the period of exponential growth lasted approximately 18 days.

```{r R0-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + # Create a ggplot
  # Choropleth map of basic reproductive number
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = R),
          color = NA) +
  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_scico(name = expression(R[0]),
                       direction = 1,
                   palette = "lajolla") +
  # Overlay counties with missing values of the reproductive number in white
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   R == 0), 
          fill = "white",
          color = NA) +
  # Overlay county boundaries
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          size = 0.05,
          fill = NA,
          color = "gray60") +
  # Add boundary of continental US
  geom_sf(data = us_geo,
          fill = NA) + 
  # Add caption
  labs(caption = "Note: counties in white represent missing values of the basic reproductive number") +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

Table \ref{tab:swn-results} reproduces the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SWN). It is possible to verify that the results match, with only the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is changed to one percent units, instead of the ten percent units used by SWN). The mixed linear model gives random intercepts (i.e., the intercept is a random variable), and the standard deviation is reported in the fifth row of Table \ref{tab:swn-results}. It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tend to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tend to have higher values of $R_0$. The key of the analysis, after extensive sensitivity analysis, is a robust finding that population density has a positive association with the basic reproductive number. But does it?

@@

```{r tabulate-swn-results, echo=FALSE}
results_swn %>%
  mutate(Variable = c("Intercept", 
                      "Log of population density",
                      "Percent of private transportation", 
                      "Median household income ($10,000)",
                      "Standard deviation (Intercept)",
                      "Within-group standard error")) %>%
  kable("html",
        digits = 3,
        booktabs = TRUE,
        col.names = c("Variable",
                      "beta",
                      "95% CI",
                      "beta",
                      "95% CI",
                      "beta",
                      "95% CI"),
        caption = "\\label{tab:swn-results}Reproducing SWN: Models 1-3") %>%
  kable_styling(latex_options = c("scale_down")) %>%
  add_header_above(c(" ", "Model 1" = 2, "Model 2" = 2, "Model 3" = 2))
```

```{r random-terms-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + 
  geom_sf(data = state_geo %>%
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = random_intercepts), # Add the simple features object to the plot
          color = "lightgray") +
  # Add boundary of continental US
  geom_sf(data = us_geo,
          fill = NA) + 
  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_scico(name = "Random Intercepts",
                   direction = -1,
                   palette = "vikO") +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r urban-areas-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:urban-areas-map}Urban areas with population > 50,000 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + # Create a ggplot
  # Choropleth map of basic reproductive number
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = R),
          color = NA) +
  # Use viridis color palette: scico color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_scico(name = "R0",
                   palette = "lajolla") +
  # Overlay counties with missing values of the reproductive number in white
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   R == 0), 
          fill = "white",
          color = NA) +
  # Overlay boundaries of urban areas
  geom_sf(data = urban_geo %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   UATYP10 == "U"), # Include only urbanized areas with population > 50,000
          fill = NA,
          color = "blue",
          size = 0.05) +
  # Add boundary of continental US
  geom_sf(data = us_geo,
          fill = NA) + 
  # Add caption
  labs(caption = "Note: boundaries of urbanized areas with population > 50,000 are shown in blue") +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r tabulate-sample-selection-results, echo=FALSE}
results_selection  %>%
  mutate(Variable = c("Intercept",
                      "Log of population density",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (10,000)",
                      "Intercept",
                      "Log of population density",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (\\$10,000)",
                      "$\\sigma$",
                      "$\\rho$")) %>%
  kable("html",
        digits = 3,
        booktabs = TRUE,
        escape = FALSE,
        col.names = c("Variable",
                      "$\\beta$", 
                      "95\\% CI",
                      "$\\beta$",
                      "95\\% CI"),
        align = c("lcccc"),
        caption = "\\label{tab:selection-results}Estimation results of sample selection models") %>%
  kable_styling(latex_options = c("scale_down")) %>% 
  pack_rows(group_label = "Sample Selection Model", 1, 6) %>%
  pack_rows(group_label = "Outcome Model", 7, 12) %>%
  add_header_above(c(" ", "Selection Model 1" = 2, "Selection Model 2" = 2))
```

@@

```{r tabulate-sample-selection3-results, echo=FALSE}
selection3_me  %>%
  mutate(Variable = c("Intercept",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (10,000)",
                      "Intercept",
                      "Density (1,000 per sq.km)",
                      "Density squared",
                      "Percent of private transportation", 
                      "Median household income (\\$10,000)",
                      "Spatial filter",
                      "$\\sigma$",
                      "$\\rho$")) %>%
  kable("html",
        digits = 3,
        booktabs = TRUE,
        escape = FALSE,
        col.names = c("Variable",
                      "$\\beta$", 
                      "95\\% CI"),
        align = c("lcccc"),
        caption = "\\label{tab:selection3-results}Estimation results of sample selection model with spatial filter") %>%
  kable_styling(font_size = 8, latex_options = "hold_position") %>% 
  pack_rows(group_label = "Sample Selection Model", 1, 5) %>%
  pack_rows(group_label = "Outcome Model", 6, 11) %>%
  add_header_above(c(" ", "Selection Model 3" = 2))
```

Clearly, the various models display some intriguing differences; but how relevant are said differences from a more substantive standpoint? Figure \ref{fig:comparison-results} shows the relationship between density and $R_0$ implied by SWN Model 3, Selection Model 2, and Selection Model 3. The left panel of the figure shows the non-linear but monotonic relationship implied by SWN Model 1. The conclusion is that at higher densities, $R_0$ is _always_ higher. The two panels on the right, in contrast, shows that Selection Model 2 and Selection Model 3 coincide that $R_0$ tends to increase as density grows. This continues until a density of approximately 2.9 (1,000 people per sq.km). At higher densities than that the relationship between density and $R_0$ begins to weaken, and the relationship becomes negative at densities higher than approximately 5.7 (1,000 people per sq.km).

@@

It is worth at this point to recall Cressie's dictum about modelling: "[w]hat is one person's mean structure could be another person's correlation structure" [@Cressie1989geostatistics, p. 201]. There are almost always multiple ways to approach a modelling situation, as lively illustrated by a recent paper that reports the results of a crowdsourced modelling experiment [@Schweinsberg2021same]. In the present case, we would argue that spatial sampling is an important aspect of the modeling process. Importantly, by adopting high reproducibility standards, SWN made a valuable contribution to the collective enterprise of seeking knowledge. Their effort, and subsequent efforts to validate and expand on their work, can potentially contribute to provide clarity to ongoing conversations about the relevance of density and the spread of COVID-19.

In particular, it is noteworthy that a sample selection model with a different variable transformation does not lend support to the thesis that higher density is _always_ associated with a greater risk of spread of the virus [in Wong and Li's words, "'Density is destiny' is probably an overstatement"; -@Wong2020spreading]. At the same time, the results presented here also stand in contrast to the findings of Hamidi et al., who found that higher density was either not significantly associated with the rate of the virus in a cross-sectional study [@Hamidi2020density], or was negatively associated with it in a longitudinal setting [@Hamidi2020longitudinal. In this sense, the conclusion that density does not aggravate the pandemic may have been somewhat premature; instead, reanalysis of the data of SWN suggests that Fielding-Miller et al. [-@Fielding2020social] might be onto something with respect to the difference between rural and urban counties. More generally, there is no doubt that in population-level studies density is indicative of proximity, but it also potentially is a proxy for adaptive behavior. And it is possible that the determining factor during COVID-19, at least in the US, has been variations in perceptions of the risks associated with contagion [@Chauhan2021covid], and subsequent compensations in behavior in more and less dense regions.

@@

Taking one recent example from the literature [Sy et al., @Sy2021population; SWN], the present paper illustrates the importance of good reproducibility practices. Sharing data and code can catalyze research, by allowing independent verification of findings, as well as additional research. After verifying the results of SWN, experiments with sample selection models and variations in the definition of model inputs, lead to an important reappraisal of the conclusion that high density is associated with greater spread of the virus. Instead, the possibility of a non-monotonical relationship between population density and contagion is raised. I do not claim that the analysis presented here is the last word on the topic of density and the spread of COVID-19, and there is always the possibility that someone else will be better equipped to analyze these data with greater competence. By opening up the analysis, documenting the way data were pre-processed, and by sharing analysis ready data, my hope would be that others will be able to discover the limitations of my own analysis and improve on it, as appropriate.

More generally, my hope is that the research of Sy et al. [-@Sy2021population], the present paper, and similar reproducible publications, will continue to encourage others to adopt higher reproducibility standards in their research.
